Disclosing funding sources for open access publication fees: the Open APC

### 2016-06-03 article.Rmd
adding springer nature analysis and remove tables and figures from the manuscript

This paper focuses on how much German universities and research organization spent on open access publication fees. In Germany, the Deutsche Forschungsgemeinschaft (DFG), the largest German research funder, has strongly influenced how universities manage institutional support for publication fees.[^3] Before the DFG started to pay for centrally funded publication fees on a pro rata basis through its "Open-Access Publishing" program  in 2011, and similar to the situation described in Canada [@Hampson_2014] or the UK [@PINFIELD_2012], only few central funds existed [@eppelin_2012]. Because the DFG has enforced a set of criteria grantees have to comply with, similar implementations for supporting open access publishing across German universities exists [@Fournier_2013]: these criteria exclude sponsoring of articles in hybrid journals, and the funding of articles whose publication fee exceeds 2,000 € (excluding value added tax). Grantees agree not only to reimburse the bills on behalf of the researchers they support, but also to look for ways to improve the handling of those financial transactions. They include central invoicing schemes and related agreements between university libraries and publishers [@Fournier_2013].

@@

# fix plos name change in journal titles
my_apc$journal_full_title <- gsub("PLoS", "PLOS", my_apc$journal_full_title)
my_apc$journal_full_title <- gsub("PLOS ONE", "PLOS ONE", my_apc$journal_full_title)

@@

On May 13, 2016,[^16a] the Open APC initiative covered `r nrow(my_apc)` articles whose publication fees were centrally paid by `r length(unique(my_apc$institution))` German universities and research institutions until 2015. The number of supported open access journal articles grew over the years (see Figure 1). While one institution disclosed `r filter(my_apc, period == min(my_apc$period)) %>% nrow()` payments made in `r as.character(min(my_apc$period))`, the majority shared their expenditures from 2013 onwards. With `r max(apc_time$articles)` articles, the year `r apc_time[apc_time$articles == max(apc_time$articles),"period"]` was best represented in our data-set. However, `r my_apc %>% filter(period == 2015) %>% group_by(institution) %>% summarise(n()) %>% nrow()` institutions contributed their cost data for 2015 at the time of this analysis, suggesting that there exits a time lag between payments made and reporting these spending to the Open APC initiative.

```{r articles_year, fig.width=6, fig.height=2.5,fig.cap="Growth of Open APC Initiative"}
apc_time <- my_apc %>%
  group_by(period) %>%
  summarize(articles = n())

## http://www.noamross.net/blog/2013/11/20/formatting-plots-for-pubs.html
science_theme = theme(panel.grid.major = element_line(size = 0.5, color = "grey"), 
    axis.line.x = element_line(size = 0.7, color = "black"),
    axis.line.y = element_line(size = 0.7, color = "black"),
    text = element_text(size = 14))

ggplot(apc_time, aes(factor(period), cumsum(articles), group = 1)) +
  geom_line(stat = "identity", colour = "blue") +
  geom_point() +
  xlab("Year") +
  ylab("OA articles reported\n(cumulative sum)") +
  scale_y_continuous(limits = c(0,8000)) +
  theme_bw() + science_theme
ggsave("figure_1.pdf", height=4, width=6)

@@

```{r histm, fig.width=6, fig.height=2.5,fig.cap="Histogram APC"}
require(ggplot2)
## http://www.noamross.net/blog/2013/11/20/formatting-plots-for-pubs.html
science_theme = theme(panel.grid.major = element_line(size = 0.5, color = "grey"), 
    axis.line.x = element_line(size = 0.7, color = "black"),
    axis.line.y = element_line(size = 0.7, color = "black"),
    text = element_text(size = 14))

ggplot(my_apc, aes(euro)) + 
  geom_histogram(binwidth = 250, colour ="black", boundary = 0) + 
  scale_x_continuous(limits = c(0,8000)) +
  xlab("Publication fee in Euro") +
  ylab("Articles") +
  theme_bw() + science_theme

@@

Among all articles, fees amounted to `r sum(my_apc$euro)` € including VAT, the average payment was `r mean(my_apc$euro)` € and the median value `r median(my_apc$euro)` € . Figure 2 shows the large price variation among the articles. The disclosed publication fees ranged from `r min(my_apc$euro)` € to `r max(my_apc$euro)` €. However, the average price paid varied somewhat during the period 2011 and 2014 (`r my_apc %>% filter(period == 2011) %>% summarise(mean(euro)) %>% round()` - `r my_apc %>% filter(period == 2014) %>% summarise(mean(euro)) %>% round()` €). We also observe that `r nrow(my_apc[my_apc$euro < 2000,])` (`r (nrow(my_apc[my_apc$euro < 2000,]) / nrow(my_apc))  *100`%) of the publication fees were paid in accordance with the DFG price cap of 2,000 €.

```{r payments_year, fig.width=6, fig.height=3,fig.cap="Payments per year"}
ggplot(my_apc, aes(factor(period), euro)) +
  geom_boxplot(outlier.size = 0.5) +
  xlab("Year") +
  ylab("Publication fee paid in EURO") +
  scale_y_continuous(limits = c(0,8000)) + 
  theme_bw() + science_theme
ggsave("figure_2.pdf", height=4, width=6)

The number of APC payments per institutions varied considerably (see Table 3). With `r filter(my_apc, institution == "MPG") %>% nrow()` reported articles, the Max Planck Society contributed `r (filter(my_apc, institution == "MPG") %>% nrow() / nrow(my_apc)) * 100` % of the overall submissions. In contrast, the two universities of technology, TU Clausthal and TU Ilmenau, who recently begun to set up support structures for fee-based open access journal articles, shared payments made for four articles each.

```{r results='asis'}
tt <- my_apc %>%
  group_by(institution) %>%
  summarise(Articles = n(), Total = sum(euro), Mean = mean(euro), Median = median(euro), Min = min(euro), Max = max(euro)) %>%
  arrange(desc(Articles))

colnames(tt) <- c("Institutions", "Articles funded", "Total", "Mean", "Median", "Minimum", "Maximum")
tt <- format(tt,big.mark=',')
pander::pandoc.table(tt, split.cells = 7, caption = 'Institutions self-reporting expenditures for open access publications (in €)', justify = 'lrrrrrr')
```

@@

```{r results='asis'}
tt <- read.csv("data/aggregates_fees_paid.csv", header = TRUE, sep =",")
colnames(tt) <- c("Cost data-set", "Journal Type", "Articles funded", "Total costs in Euro", "Mean")
pander::pandoc.table(tt,split.cells = 5 , big.mark = ",", caption = 'Comparision of cost data per period and journal type  (in €)', justify = 'lrrrr')
```

@@

Along with the price information, participating institutions were required to identify funded articles by their DOI. They were reported for `r my_apc %>% filter(!is.na(doi)) %>% nrow()` out of `r nrow(my_apc)` articles. Of those, `my_apc %>% filter(indexed_in_crossref == TRUE) %>% nrow()` were indexed in CrossRef, representing `my_apc %>% filter(indexed_in_crossref == TRUE) %>% nrow() / my_apc %>% nrow() * 100`  % of all funded publications. The reasons why articles identified by a DOI were not registered with CrossRef differed. Some journals were not indexed by CrossRef at the time of our study but by the DOI agencies DataCite (Journal of new frontiers in spatial concepts published by KIT Scientific Publishing) and Medra (DIE ERDE: Journal of the Geographical Society of Berlin). In other cases, either the DOI did not refer to the full text despite the fact that the journal was indexed on a regular basis (compare <http://doi.org/10.1186/1471-2105-13-S19-S7> with <http://bmcbioinformatics.biomedcentral.com/articles/10.1186/1471-2105-13-S19-S7>) or the resource type could not be retrieved, although the DOI resolves (<http://doi.org/10.1186/s12885-015-1795-7>).

@@

@@

colnames(tt) <- c("Journal", "Articles funded", "Total", "Mean", "SD", "Median", "Range")
tt <- format(tt,big.mark=',')
pander::pandoc.table(tt, split.cells = 7, caption = 'Publication fees paid per publisher (in €)', justify = 'lrrrrrr')
```

Most of the funding of publication fees in Germany went to the publisher Springer Science + Business Media, especially profiting from the merge with the former full open access publisher BioMed Central. In contrast, other established publishing houses such as Elsevier and Wiley-Blackwell rank lower, presumably because they mostly publish hybrid journals, which were not well represented in our data-set at the time of the study. Table 4 also illustrates the variation across and within publishers, which confirms earlier findings [@Pinfield_2015].

@@

colnames(tt) <- c("Journal", "Articles funded", "Total", "Mean", "SD", "Median", "Range")
tt <- format(tt,big.mark=',')
pander::pandoc.table(tt, split.cells = 7, caption = 'Publication fees paid per journal (in €)', justify = 'lrrrrrr')
```

>>>>

This paper focuses on how much German universities and research organization spent on open access publication fees. In Germany, the Deutsche Forschungsgemeinschaft (DFG), the largest German research funder, has strongly influenced how universities manage institutional support for publication fees.[^3] Before the DFG started to pay for centrally funded publication fees on a pro rata basis through its "Open-Access Publishing" program  in 2011, and similar to the situation described in Canada [@Hampson_2014] or the UK [@PINFIELD_2012], only few central funds existed [@eppelin_2012]. Because the DFG has enforced a set of criteria grantees have to comply with, similar implementations for supporting open access publishing across German universities exists [@Fournier_2013]. These criteria exclude sponsoring of articles in hybrid journals, and the funding of articles whose publication fee exceeds 2,000 € (excluding value added tax)[^3]. Grantees agree not only to reimburse the bills on behalf of the researchers they support, but also to look for ways to improve the handling of those financial transactions. They include central invoicing schemes and related agreements between university libraries and publishers [@Fournier_2013].

@@

# fix plos name change in journal titles
my_apc$journal_full_title <- gsub("PLoS", "PLOS", my_apc$journal_full_title)
my_apc$journal_full_title <- gsub("Scientific Repor.", "Scientific Reports", my_apc$journal_full_title)

@@

After excluding payments being made for non-journal articles and in 2016, we retrieved `r nrow(my_apc)` articles for which the publication fee was centrally paid by `r length(unique(my_apc$institution))` German universities and research institutions until 2015. The number of supported open access journal articles grew over the years (see Figure 1). While one institution disclosed `r filter(my_apc, period == min(my_apc$period)) %>% nrow()` payments made in `r as.character(min(my_apc$period))`, the majority shared their expenditures from 2013 onwards. With `r max(apc_time$articles)` articles, the year `r apc_time[apc_time$articles == max(apc_time$articles),"period"]` was best represented in our data-set. However, `r my_apc %>% filter(period == 2015) %>% group_by(institution) %>% summarise(n()) %>% nrow()` institutions contributed their cost data for 2015 at the time of this analysis, suggesting that there exits a time lag between payments made and reporting these spending to the Open APC initiative.

**Figure 1: Growth of Open APC Initiative**

```{r articles_year, results = 'hide'}
apc_time <- my_apc %>%
  group_by(period) %>%
  summarize(articles = n())

## http://www.noamross.net/blog/2013/11/20/formatting-plots-for-pubs.html
science_theme = theme(panel.grid.major = element_line(size = 0.5, color = "grey"), 
    axis.line.x = element_line(size = 0.7, color = "black"),
    axis.line.y = element_line(size = 0.7, color = "black"),
    text = element_text(size = 14))
pdf("figure_1.pdf", height=4, width=6) 
ggplot(apc_time, aes(factor(period), cumsum(articles), group = 1)) +
  geom_line(stat = "identity", colour = "blue") +
  geom_point() +
  xlab("Year") +
  ylab("OA articles reported\n(cumulative sum)") +
  scale_y_continuous(limits = c(0,8000)) +
  theme_bw() + science_theme
dev.off()

@@

```{r histm, results = 'hide'}
require(ggplot2)
## http://www.noamross.net/blog/2013/11/20/formatting-plots-for-pubs.html
science_theme = theme(panel.grid.major = element_line(size = 0.5, color = "grey"), 
    axis.line.x = element_line(size = 0.7, color = "black"),
    axis.line.y = element_line(size = 0.7, color = "black"),
    text = element_text(size = 14))
pdf("figure_2.pdf", height=4, width=6)
ggplot(my_apc, aes(euro)) + 
  geom_histogram(binwidth = 250, colour ="black", boundary = 0) + 
  scale_x_continuous(limits = c(0,8000)) +
  xlab("Publication fee in Euro") +
  ylab("Articles") +
  theme_bw() + science_theme
dev.off()
```

Among all articles, fees amounted to `r sum(my_apc$euro)` € including VAT, the average payment was `r mean(my_apc$euro)` € and the median value `r median(my_apc$euro)` € .

**Figure 2: Distribution of institutional spending on publication fees by German research organisations**

**Figure 3: Institutional spending on publication fees by German research organisations per year**

Figure 3 presents the large price variation among the articles. The disclosed publication fees ranged from `r min(my_apc$euro)` € to `r max(my_apc$euro)` €. However, the average price paid varied somewhat during the period 2011 and 2014 (`r my_apc %>% filter(period == 2011) %>% summarise(mean(euro)) %>% round()` - `r my_apc %>% filter(period == 2014) %>% summarise(mean(euro)) %>% round()` €). We also observe that `r nrow(my_apc[my_apc$euro < 2000,])` (`r (nrow(my_apc[my_apc$euro < 2000,]) / nrow(my_apc))  *100`%) of the publication fees were paid in accordance with the DFG price cap of 2,000 €.

```{r payments_year, results = 'hide'}
pdf("figure_3.pdf", height=4, width=6)
ggplot(my_apc, aes(factor(period), euro)) +
  geom_boxplot(outlier.size = 0.5) +
  xlab("Year") +
  ylab("Publication fee paid in EURO") +
  scale_y_continuous(limits = c(0,8000)) + 
  theme_bw() + science_theme
dev.off()
```

The number of APC payments per institutions varied considerably (see Table 1). With `r filter(my_apc, institution == "MPG") %>% nrow()` reported articles, the Max Planck Society contributed `r (filter(my_apc, institution == "MPG") %>% nrow() / nrow(my_apc)) * 100` % of the overall submissions. In contrast, the  universities of technology, TU Clausthal, TU Dortmund TU Ilmenau that recently begun to set up support structures for fee-based open access journal articles, self-reported low numbers of payments.

**Table 1: Institutional spending on open access publications (in €)**

```{r}
tt <- my_apc %>%
  group_by(institution) %>%
  summarise(Articles = n(), Total = sum(euro), Mean = mean(euro), SD = sd(euro), Median = median(euro), Min = min(euro), Max = max(euro)) %>%
  mutate(range_apc = paste(Min, "--", Max)) %>%
  select(-Min, -Max) %>%
  arrange(desc(Articles))

colnames(tt) <- c("Institution", "Articles funded", "Total", "Mean", "SD", "Median", "Min-Max")
table_1 <- format(tt,big.mark=',')
#table_1 <- pander::pandoc.table(tt, split.cells = 7, caption = 'Publication fees paid per publisher (in €)', justify = 'lrrrrrr')

@@

@@

```{r results='hide'}
tt <- read.csv("data/aggregates_fees_paid.csv", header = TRUE, sep =",")
colnames(tt) <- c("Cost data-set", "Journal Type", "Articles funded", "Total costs in Euro", "Mean")
table_2 <- tt
#table_2 <- pander::pandoc.table(tt,split.cells = 5 , big.mark = ",", caption = 'Comparision of cost data per period and journal type  (in €)', justify = 'lrrrr')
```

@@

```{r}
doi_miss <- jsonlite::stream_in(file("data/missing_dois_agency.json"), verbose = FALSE)
doi_miss[doi_miss$doi_agency == "", "doi_ageny"] <- NA
doi_miss <- doi_miss[!doi_miss$institution == "FWF - Austrian Science Fund",]

cr <- my_apc %>% 
  filter(!doi %in% doi_miss$doi) %>%
  filter(doi = !is.na(doi))

agencies <- doi_miss %>% 
  group_by(doi_agency) %>% 
  summarise(n()) %>%
  rbind(c("CrossRef", nrow(cr)),.)

agencies <- data.frame(table(unlist(doi_miss$doi_agency)))

To identify the sponsored open access journal articles, as well as to gather bibliographic metadata, DOIs were a mandatory part of the Open APC initiative's data collection activities. The participating institutions reported DOIs for `r my_apc %>% filter(!is.na(doi)) %>% nrow()`  out of `r nrow(my_apc)` articles. using these DOIs, we retrieved additional metadata from CrossRef for `r nrow(cr)` publications,  representing `r nrow(cr) / nrow(my_apc) * 100`  % of the total article volume. Articles for which no metadata could be obtained, were either registered with the DOI agency DataCite (10 articles) or Medra (two articles). For eight articles, our parser could not gather the XML resource, although these publications were registered with CrossRef. Seven DOIs reported to the Open APC initiative did not resolve.

@@

**Table 2: Publication fees paid per publisher (in €)**

@@

colnames(tt) <- c("Journal", "Articles funded", "Total", "Mean", "SD", "Median", "Range")
table_2 <- format(tt,big.mark=',')
# pander::pandoc.table(tt, split.cells = 7, caption = 'Publication fees paid per publisher (in €)', justify = 'lrrrrrr')

## springer nature merger
springer_nat <- my_apc %>% 
  filter(publisher == "Springer Nature")

springer <- springer_nat %>% filter(member_id == "297")
nature <- springer_nat %>% filter(member_id == "339")
```

Most of the funding of publication fees in Germany went to the publisher Springer Nature, most llikeky profiting from the merge with the open access publisher BioMed Central in 2008, and in 2015 between the well-established publishers Springer Science + Business Media and Nature Publishing Group. Using the CrossRef-Member-ID instead of the publisher name, we were able to differentiate between journals  formely published by Springer Science + Business Media and Nature Publishing Group. In terms of articles, the majority of payments made were for publications in journal that were formerly associated with Springer Science + Business Media. Springer Science + Business Media journals accounted for `r nrow(springer)`articles, representing `r nrow(springer) / nrow(springer_nat) * 100` of the total Springer Nature article volume in the spending data, and `r sum(springer$euro) / sum(springer_nat$euro) * 100` % of the amount spent. Median publication fee spending slightly differed between Springer Science + Business Media (`r median(springer$euro)` €) and Nature Publishing (`r median(nature$euro)` €). However, price variation was higher for Nature Publishing journals (SD = `r sd(nature$euro)`€) compared with that of the former Springer Science + Business Media titles (SD = `r sd(nature$euro)` €).

In contrast to Springer Nature, other well-established publishing houses such as Elsevier and Wiley-Blackwell rank lower in our analysis, presumably because they mostly published hybrid journals, a journal type that was not well represented in our data-set at the time of this study. Table 4 also illustrates the variation across and within publishers, which confirms earlier findings [@Solomon_2012; @Pinfield_2015].

**Table 3: Publication fees paid per journal (in €)**

@@

colnames(tt) <- c("Journal", "Articles funded", "Total", "Mean", "SD", "Median", "Range")
table_3 <- format(tt,big.mark=',')
#pander::pandoc.table(tt, split.cells = 7, caption = 'Publication fees paid per journal (in €)', justify = 'lrrrrrr')
```
